Are we learning to build better models at the cost of better decisions?

### 2021-03-20 manuscript/appendix_B.Rmd
:pen:

```{r}
rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
Ks <- c(0.5, 1, 1.5, 4, 10, 16)
n_models <- length(rs)*length(Ks)
sigmas <- 0.1
pars <- expand.grid(rs, Ks)
names(pars) <- c("r", "K")
pars$sigma <- sigmas

@@

On a discrete grid of possible states and actions, we can define the growth rate of a given state $X_t$ subject to harvest $H_t$,
$f(X_t,H_t)$ as set of matrices.  Each matrix $i$ gives the transition probabilities for any current state to any future state, 
given that action $i$ is taken.

@@

@@

i <- 10:20
r_marginals <- posteriors %>% group_by(time,r) %>% 
  summarise(prob = sum(probability), .groups = "drop") %>% 
  filter(time %in% i)

r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_line() + facet_wrap(~time)
```

@@

```{r}
i = 19
K_marginals <- posteriors %>% group_by(time,K) %>% 
  summarise(prob = sum(probability), .groups = "drop") %>% 
  filter(time %in% i) 
K_marginals %>%
  ggplot(aes(K, prob, group=time)) + geom_path() + geom_point() + facet_wrap(~time)

@@

```{r}
Tmax <- max(am_many$df$time)
best_i <- which.max(am_many$posterior[Tmax,])
good_i <- am_many$posterior[Tmax,] > 0.1
pars[good_i,] %>% mutate(final_prob = as.numeric(am_many$posterior[Tmax,good_i])) %>% knitr::kable("latex")
```

```{r, fig.cap = ""}
model1 <- dim(am_many$posterior)[2]
am_many$df %>% 
  mutate(belief = am_many$posterior[,model1]) %>%
  ggplot(aes(time, states[state], col = belief)) + 
  geom_line() + 
  geom_point() +
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")

>>>>

```{r}
# Try 6 log-spaced values
rs <- 10 ^ seq(0.01, 2, length=6)
Ks <- 10 ^ seq(-.3, 1.2, length=6)

# compare over fixed set of candidate r and K
rs <- c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 2)
Ks <- c(0.5, 1, 1.5, 4, 10, 16)
n_models <- length(rs)*length(Ks)
sigmas <- 0.1
pars <- expand.grid(rs, Ks)
names(pars) <- c("r", "K")
pars$sigma <- sigmas

@@

@@

```{r}
Tmax <- max(am_many$df$time)
best_i <- which.max(am_many$posterior[Tmax,])
good_i <- am_many$posterior[Tmax,] > 0.1
pars[good_i,] %>% mutate(final_prob = as.numeric(am_many$posterior[Tmax,good_i])) 
```

```{r}
model1 <- dim(am_many$posterior)[2]

am_multi <- am_many$df %>% 
  mutate(belief = am_many$posterior[,model1],
         stock = states[state],
         quota = actions[action]) %>%
  select(time, belief, stock, quota) 
write_csv(am_multi, "../data/am_multi.csv")
```

```{r}
am_multi %>%
  ggplot(aes(time, stock, col = belief)) + 
  geom_line() + 
  geom_point() +
  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")

@@

i <- 20
r_marginals <- posteriors %>% group_by(time,r) %>% 
  summarise(prob = sum(probability), .groups = "drop") %>% 
  filter(time %in% i)

r_marginals %>%  ggplot(aes(r, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time)
```

@@

```{r}
i = 20
K_marginals <- posteriors %>% group_by(time,K) %>% 
  summarise(prob = sum(probability), .groups = "drop") %>% 
  filter(time %in% i) 
K_marginals %>%
  ggplot(aes(K, prob, group=time)) + geom_bar(stat="identity") + facet_wrap(~time) + scale_x_log10()

@@

```{r}
policy <- mdp_compute_policy(transitions[-1],  
                           reward = reward, 
                           discount = discount,
                           model_prior = model_prior,
                           Tmax = 20)
non_adaptive <- memoise::memoise(mdp_planning, cache = memoise::cache_filesystem("cache/"))
x0 <- which.min(abs(states - 6))
## start with 90% weight on 'model 1'
model_prior <- c(rep(.1, length(models)-2) / (length(models) - 2), .9)

non_am <- non_adaptive(transitions[-1],  
                           reward = reward, 
                           discount = discount,
                           model_prior = model_prior,
                           x0 = x0, 
                           policy = policy,
                           Tmax = 20, 
                           true_transition = transitions[[true_model]], 
                           epsilon = 0.001, 
                           max_iter = 2000)
```

### 2021-03-20 manuscript/manuscript.Rmd
:pen:

## change to true to add optional line numbering
lineno: true

output: rticles::elsevier_article
layout: 3p
---

@@

Whenever models are estimated from the data they seek to describe, there is a concern that
the model may over-fit the data: making the data seem even more likely than it would be
under the true process [@Burnham2002].  Over-fitting is a central concern of any model comparison
because over-fit models may underestimate uncertainty and make poor predictions about future
outcomes [@Ginzburg2004]. By comparing outcomes predicted by a model to future observations, 
a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy.  
This iterative learning process of learning will not only decrease uncertainty by narrowing 
beliefs over plausible model or parameter values, but can also eliminate models which overfit the data.
Because any iterative learning compares models to future outcomes, overfitting those outcomes
is impossible: a concept formalized rigorously by @Gneiting2007's proof of "proper" scoring rules.
By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on
average, than that of the underlying process, $P(x)$.  @Gneiting2007 further shows which
techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which
cannot be over-fit.  The ability to decrease uncertainty while avoiding over-fitting makes
such iterative learning particularly appealing approach to decision making in an uncertain
world [@Deitze2018], and the approach is already well established in the principle of 
adaptive management [@Walters1978] and existing practices such as Management Strategy Evaluation (MSE)
[@Punt2006] used in fisheries. Given this ability to decrease uncertainty and avoid over-fitting,
such iterative updating approaches seem irreproachable from a theoretical basis, even if
they may be difficult to execute in practice.

However, as long as the true process model is unknown (effectively always), there is no
such theoretical guarantee that iterative learning over possible models or parameters will
lead to better outcomes.  This is not -- cannot -- be because some models have over-fit the 
data, as least as long as proper scoring rules are used for evaluation.  As I illustrate here,
given a set of candidate models, iterative learning either through assessment of forecast
skill or through formal adaptive management can lead to worse outcomes than forgoing any learning step.
Put another way, the value of information [@Howard1966], as measured by the expected utility given that 
information minus the utility without it, can be negative: a rational decision maker ought to pay
not to know information even when that information is perfectly accurate.  
This situation is a direct consequence of the fact that a model need only capture certain
features of the underlying process to lead to an effective decision.

@@

```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicates the belief that Model 2 is correct (red). Each model is assigned an initial prior belief. To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.  Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2."}
am <- read_csv("../data/am.csv")
am %>%  filter(time <30) %>%
  ggplot(aes(time, states[state], col = belief)) + 
  geom_line() + 
  geom_point() +
  scale_colour_gradient(limits = c(0,1), low = pal[1], high = pal[4]) +
  ylab("stock size")
```

Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve the problem [Fig 4].

@@

<!-- Not just overfitting -->
Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@overfit].  In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.  The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy. In the example here, both Model 1 and Model 2 have the same structural complexity.  The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit. Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$  or information criteria [@Burnham], will all prefer Model 2 over Model 1 for the same reason.

<!-- Not just "get more models" -->
This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2. Model 1 approximates the key feature, giving nearly optimal outcomes.  Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood []. Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations. In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ [Fig 4], or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).  Models will always be simpler than reality.  To insist that this issue can and will be avoided by always including more and better models of the process misses the point.

>>>>

## change to true to add optional line numbering
lineno: true
csl: ecology-letters.csl
output: rticles::elsevier_article
layout: 3p
journal: TBD
---

@@

Whenever models are estimated from the data they seek to describe, there is a concern that
the model may over-fit the data: making the data seem even more likely than it would be
under the true process [@Burnham1998].  Over-fitting is a central concern of any model comparison
because over-fit models may underestimate uncertainty and make poor predictions about future
outcomes [@Ginzburg2004]. By comparing outcomes predicted by a model to future observations, 
a decision maker can not only _plan_ for the uncertainty, but also _learn_ which models are most trustworthy.  
This iterative learning process of learning will not only decrease uncertainty by narrowing 
beliefs over plausible model or parameter values, but can also eliminate models which overfit the data.
Because any iterative learning compares models to future outcomes, overfitting those outcomes
is impossible: a concept formalized rigorously by @Gneiting2007's proof of "proper" scoring rules.
By definition, under a proper scoring rule, no probabilistic prediction $Q(x)$ can score better, on
average, than that of the underlying process, $P(x)$.  @Gneiting2007 further shows which
techniques for evaluating forecast predictions do an do not satisfy this condition, i.e. which
cannot be over-fit.  The ability to decrease uncertainty while avoiding over-fitting makes
such iterative learning particularly appealing approach to decision making in an uncertain
world [@Dietze2018], and the approach is already well established in the principle of 
adaptive management [@Walters1978] and existing practices such as Management Strategy Evaluation (MSE)
[@Punt2016] used in fisheries. Given this ability to decrease uncertainty and avoid over-fitting,
such iterative updating approaches seem irreproachable from a theoretical basis, even if
they may be difficult to execute in practice.

However, as long as the true process model is unknown (effectively always), there is no
such theoretical guarantee that iterative learning over possible models or parameters will
lead to better outcomes.  This is not -- cannot -- be because some models have over-fit the 
data, as least as long as proper scoring rules are used for evaluation.  As I illustrate here,
given a set of candidate models, iterative learning either through assessment of forecast
skill or through formal adaptive management can lead to worse outcomes than forgoing any learning step.
Put another way, the value of information [VOI; @Howard1966; @Katz1987], as measured by the expected utility given that 
information minus the utility without it, can be negative: a rational decision maker ought to pay
not to know information even when that information is perfectly accurate.  
This situation is a direct consequence of the fact that a model need only capture certain
features of the underlying process to lead to an effective decision.

@@

```{r figure3, fig.cap="Adaptive management under model uncertainty, color indicates the belief that Model 1 is correct (blue). Each model is assigned an initial prior belief. To ensure that any issues with passive exploration are not the cause of preferring Model 2, the initial belief in Model 2 is set to 1%.  Within a single iteration of adaptive management, the belief over models is updated to near certainty in Model 2, resulting in higher harvests and lower stock sizes similar to managing under Model 2 alone, Fig 2."}
am <- read_csv("../data/am.csv")
am %>%  filter(time <=20) %>%
  ggplot(aes(time, stock, col = belief)) + 
  geom_line() + 
  geom_point() +
  geom_point(aes(time, quota), col="darkgreen", shape=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) +
  ylab("stock size")
```

Increasing the space of possible models to cover a whole plausible range of parameters $r$ and $K$ does nothing to resolve this problem.  Learning is somewhat slower over the larger range of possibilities, but nevertheless converges towards parameter values with lower growth rates and lower carrying capacity, for which the target biomass is far lower [Fig 4]. The manager becomes trapped in vicious cycle as biomass declines further and further, subsequent observations seem more and more consistent with models which most favor over-fishing in the first place.  As biomass declines, annual harvests and profits also plunge, and yet our manager becomes only more confident in their decisions.

```{r}
am_multi <- read_csv("../data/am_multi.csv")
am_multi %>%
  ggplot(aes(time, stock, col = belief)) + 
  geom_line() + 
  geom_point() +
  geom_point(aes(time, harvest), col="darkgreen", shape=2) + 
  scale_colour_gradient(limits = c(0,1), low = pal[4], high = pal[1]) + 
  ylab("fish stock") + labs(caption = "Belief in model 1 falls off rapidly as magagement pushes stocks to lower and lower sizes.")
```

@@

<!-- Not just overfitting -->
Note that the mechanism shown here has nothing to do with the much more familiar issue of over-fitting, in which a better-fitting model will also lead to worse outcomes [@Burnham1998].  In fact, an model which has been over-fit will gradually be rejected by either the iterative forecasting or adaptive management approaches shown here, as these approaches continually confront the models with new data to which they had not been previously fit.  The ability to avoid over-fitting is one of the greatest appeals of any iterative management strategy. In the example here, both Model 1 and Model 2 have the same structural complexity.  The better decision performance of Model 1 is a consequence of capturing the key attribute needed for a good decision, which is perhaps a surprisingly an all-together different criteria than fit. Note that other methods of model selection not considered here, including goodness-of-fit metrics such as $r^2$  or information criteria [@Burnham1998], will all prefer Model 2 over Model 1 for the same reason.

<!-- Not just "get more models" -->
This case cannot be dismissed merely as being dealt a bad hand in having to pick only between Model 1 and Model 2. Model 1 approximates the key feature, giving nearly optimal outcomes.  Previous literature has underscored the importance of Knightian "unknown unknowns" two alternative models examined here fail to fully reflect our uncertainty in the underlying dynamics, and the consequences of underestimating model uncertainty are well understood [@Polasky2011; @Wintle2010]. Had we included the true model in the set of possibilities, the techniques illustrated would have had little difficulty in distinguishing it from Models 1 and 2 after sufficient iterations. In practice, we never have the true model, and even a much more thorough suite of candidate models, such as considering any possible values of $r$, $K$ [Fig 4], or even among common alternative models with different structural forms, such as Beverton-Holt [@Beverton1957], Ricker [@Ricker1954], or Shepherd [@Shepherd1980], would fail to do any better, since all these models still have symmetric growth rates with a peak growth rate at half the steady-state population size).  Models will always be simpler than reality.  To insist that this issue can and will be avoided by always including more and better models of the process misses the point.
